The Poisson approximate likelihood compared to the particle filter

### 2024-07-24 ms.Rnw
compile

\begin{document}
\SweaveOpts{concordance=TRUE}

@@

RUN_LEVEL = 1
NP_MIF       = switch(RUN_LEVEL, 5000, 50000)
NMIF         = switch(RUN_LEVEL, 4,    100)
ncores       = switch(RUN_LEVEL, 2,    36)
top_n_fits   = switch(RUN_LEVEL, 1,    12)
sim_rep      = switch(RUN_LEVEL, 1,    100) # Total number of simulations

@@

<<OvOv-model, echo=False, massage=False, warning=False, results=hide>>=
# measurement model 
dmeas <- Csnippet("
                  if (ISNA(cases1)) {
                  lik = (give_log) ? 0 : 1;
                  } else {
                        lik =  dbinom(cases1, H1, q1, 1) +
                        dbinom(cases2, H2, q2, 1) +
                        dbinom(cases3, H3, q3, 1);

@@

<<maximization1, echo=False, massage=False, warning=False, results=hide>>=
set.seed(123)
sir_box <- rbind(
  beta1=c(10,15),
  beta2=c(0.2,0.4),
  beta3=c(0.3,0.5),
  phi=c(0.01,0.3),
  beta11=c(0.1,0.2),
  sigma_q=c(0.001,0.1),
  sigma_xi=c(65,70),
  S10 = c(0.01,0.05),
  I10 = c(0.0001, 0.0005),
  R10 = c(0.01, 0.05),
  S20 = c(0.01,0.7),
  I20 = c(0.000001,0.00002),
  R20 = c(0.001,0.005),
  S30 = c(0.01,0.5),
  I30 = c(0.000001, 0.0001),
  R30 = c(0.0001,0.005)
)

@@

bake(file = paste0("output/round_01/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_01.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_01 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01,beta2=0.01,beta3=0.01,
                    beta11=0.01,phi=0.01,sigma_q=0.01,sigma_xi=0.01,
                    S10=ivp(0.24),I10=ivp(0.24),R10=ivp(0.24),
                    S20=ivp(0.24),I20=ivp(0.24),R20=ivp(0.24),
                    S30=ivp(0.24),I30=ivp(0.24),R30=ivp(0.24)),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting,
      block = F
    ) 
  }
  mifs_global_01
})

@@

<<maximization2, echo=False, massage=False, warning=False, results=hide>>=
set.seed(1234)
sd_02           = 2/3

@@

bake(file = paste0("output/round_02/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_02.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_02 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
                    beta11=0.01*sd_02,phi=0.01*sd_02,
                    sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
                    S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
                    S20=ivp(0.24)*sd_02,I20=ivp(0.24)*sd_02,R20=ivp(0.24)*sd_02,
                    S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting,
      block = F
    ) 
  }
  mifs_global_02
})

@@

<<maximization3, echo=False, massage=False, warning=False, results=hide>>=
set.seed(12345)
sd_03           = (2/3)^2

@@

bake(file = paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_03.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_03 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
                    beta11=0.01*sd_03,phi=0.01*sd_03,
                    sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
                    S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
                    S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
                    S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting,
      block = F
    ) 
  }
  mifs_global_03
})

@@

<<estimated-mle, echo=False, massage=False, warning=False>>=
mif_el <- readRDS(paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_03.rds")) 
# Can be changed if more rounds needed
est_mle_vec <- mif_el[which.max(mif_el$logLik),]
@

@@

<<pf-sim-100, echo=False, results=hide, massage=False, warning=False>>=
# Set WWR's MLE
wwr_mle_raw <- c(params_fixed, 
             "beta1" = 11.48,
             "beta2" = 0.25,
             "beta3" = 0.35,
             "phi" = 0.14,
             "beta11" = 0.16,
             "sigma_q" = 0.021,
             "sigma_xi" = 66.89)

@@

<<pfilter_sim_100, echo=False, massage=False, warning=False, results=hide>>=
ovov_sim_100 <- readRDS(paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"))
bake(file=paste0("output/pfilter/pfilter_ovov_sim_100_",RUN_LEVEL,".rds"),{
  pfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

<<palfilter_sim_100, echo=False, massage=False, warning=False, results=hide>>=
readRDS(paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"))
prop <- 420

@@

\begin{table}[htbp]
  \centering
  \begin{threeparttable}
    \caption{Maximum likelihood estimation of parameters by PAL for three models \citep{wwr}, and by mif2 algorithm \citep{pomppackagepaper} for the OvOv model.
    \ed{LET'S DISCUSS WHETHER WE WANT TO SHOW JUST OV-OV HERE. IT WOULD BE SIMPLER.}
    }
    \label{tab:mlebywwr}
    \begin{tabular}{>{\raggedright\arraybackslash}p{2cm} *{3}{>{\centering\arraybackslash}p{1.5cm}} >{\centering\arraybackslash}p{2.5cm}}
      \toprule
      Parameter      & EqEq  & EqOv  & OvOv  & OvOv (\textbf{pomp})$^*$ \\
      \midrule
      \(\beta_1\)    & 12.15 & 12.74 & 11.48 & $\Sexpr{round(est_mle_vec["beta1"],2)}$\\
      \(\beta_2\)    & 0.22  & 0.21  & 0.25  & $\Sexpr{round(est_mle_vec["beta2"],2)}$\\
      \(\beta_3\)    & 0.34  & 0.31  & 0.35  & $\Sexpr{round(est_mle_vec["beta3"],2)}$\\
      \(\phi\)       & 0.017 & 0.14  & 0.14  & $\Sexpr{round(est_mle_vec["phi"],2)}$\\
      \(\rho\)       & 0.022 & 0.19  & 0.16  & $\Sexpr{round(est_mle_vec["beta11"],2)}$\\
      \(\sigma^2_q\) & n/a   & 0.042 & 0.021 & $\Sexpr{round(est_mle_vec["sigma_q"],2)}$\\
      \(\sigma_\xi\) & n/a   & n/a   & 66.89 & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
      \addlinespace
      \multicolumn{1}{l}{\(S_{10}\)} & \multicolumn{3}{c}{$3876549^*$} & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$\\
      \multicolumn{1}{l}{\(I_{10}\)} & \multicolumn{3}{c}{$30351$}     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$\\
      \multicolumn{1}{l}{\(R_{10}\)} & \multicolumn{3}{c}{$1315221$}   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$\\
      \multicolumn{1}{l}{\(S_{20}\)} & \multicolumn{3}{c}{$57139612$}  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$\\
      \multicolumn{1}{l}{\(I_{20}\)} & \multicolumn{3}{c}{$871$}       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$\\
      \multicolumn{1}{l}{\(R_{20}\)} & \multicolumn{3}{c}{$302852$}    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$\\
      \multicolumn{1}{l}{\(S_{30}\)} & \multicolumn{3}{c}{$19573727$}  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$\\
      \multicolumn{1}{l}{\(I_{30}\)} & \multicolumn{3}{c}{$2550$}      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$\\
      \multicolumn{1}{l}{\(R_{30}\)} & \multicolumn{3}{c}{$131092$}    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$\\
      \midrule
      \addlinespace
      \textbf{AIC}  & $98866.65$ & $15154.75$ & $13778.08$ & $\Sexpr{round((-2)*((length(wwr_mle) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$\\
      \addlinespace
      \bottomrule
    \end{tabular}
    \begin{tablenotes}
      \small
      \item[*] The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be \textbf{fixed} in \cite{wwr}. The results of the estimates of the OvOv (\textbf{pomp}) model are obtained by using the Iterated Filtering algorithm with 3 rounds and 100 iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round. 
    \end{tablenotes}
  \end{threeparttable}
\end{table}

>>>>

\begin{document}
%\SweaveOpts{concordance=TRUE}

@@

RUN_LEVEL = 1
NP_MIF       = switch(RUN_LEVEL, 1000, 50000)
NMIF         = switch(RUN_LEVEL, 4,    100)
ncores       = switch(RUN_LEVEL, 2,    36)
top_n_fits   = switch(RUN_LEVEL, 1,    12)
sim_rep      = switch(RUN_LEVEL, 1,    100) # Total number of simulations

@@

<<OvOv-model, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
# measurement model 
dmeas <- Csnippet("
                  if (ISNA(cases1)) {
                  lik = (give_log) ? 0 : 1;
                  } else {
                        lik =  dbinom(cases1, H1, q1, 1) +
                        dbinom(cases2, H2, q2, 1) +
                        dbinom(cases3, H3, q3, 1);

@@

<<maximization1, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
set.seed(123)
sir_box <- rbind(
  beta1=c(10,15),
  beta2=c(0.2,0.4),
  beta3=c(0.3,0.5),
  phi=c(0.01,0.3),
  beta11=c(0.1,0.2),
  sigma_q=c(0.001,0.1),
  sigma_xi=c(65,70),
  S10 = c(0.01,0.05),
  I10 = c(0.0001, 0.0005),
  R10 = c(0.01, 0.05),
  S20 = c(0.01,0.7),
  I20 = c(0.000001,0.00002),
  R20 = c(0.001,0.005),
  S30 = c(0.01,0.5),
  I30 = c(0.000001, 0.0001),
  R30 = c(0.0001,0.005)
)

@@

bake(file = paste0("output/round_01/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_01.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_01 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01,beta2=0.01,beta3=0.01,
                    beta11=0.01,phi=0.01,sigma_q=0.01,sigma_xi=0.01,
                    S10=ivp(0.24),I10=ivp(0.24),R10=ivp(0.24),
                    S20=ivp(0.24),I20=ivp(0.24),R20=ivp(0.24),
                    S30=ivp(0.24),I30=ivp(0.24),R30=ivp(0.24)),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting
    ) 
  }
  mifs_global_01
})

@@

<<maximization2, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
set.seed(1234)
sd_02 = 2/3

@@

bake(file = paste0("output/round_02/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_02.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_02 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
                    beta11=0.01*sd_02,phi=0.01*sd_02,
                    sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
                    S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
                    S20=ivp(0.24)*sd_02,I20=ivp(0.24)*sd_02,R20=ivp(0.24)*sd_02,
                    S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting_values_02[[i]]
    ) 
  }
  mifs_global_02
})

@@

<<maximization3, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
set.seed(12345)
sd_03           = (2/3)^2

@@

bake(file = paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_03.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_03 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
                    beta11=0.01*sd_03,phi=0.01*sd_03,
                    sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
                    S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
                    S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
                    S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting_values_03[[i]]
    ) 
  }
  mifs_global_03
})

@@

<<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
mif_el <- readRDS(paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_03.rds")) 
# Can be changed if more rounds needed
est_mle_vec <- mif_el[which.max(mif_el$logLik),]
@

@@

<<pf-sim-100, echo=FALSE, results='hide', message=FALSE, warning=FALSE>>=
# Set WWR's MLE
wwr_mle_raw <- c(params_fixed, 
             "beta1" = 11.48,
             "beta2" = 0.25,
             "beta3" = 0.35,
             "phi" = 0.14,
             "beta11" = 0.16,
             "sigma_q" = 0.021,
             "sigma_xi" = 66.89)

@@

<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"))
bake(file=paste0("output/pfilter/pfilter_ovov_sim_100_",RUN_LEVEL,".rds"),{
  pfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

<<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
readRDS(paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"))
prop <- 420

@@

\begin{table}[htbp]
  \centering
  \begin{threeparttable}
    \caption{Maximum likelihood estimation of parameters by PAL for three models \citep{wwr}, and by mif2 algorithm \citep{pomppackagepaper} for the OvOv model.
    \ed{LET'S DISCUSS WHETHER WE WANT TO SHOW JUST OV-OV HERE. IT WOULD BE SIMPLER.}
    }
    \label{tab:mlebywwr}
    \begin{tabular}{>{\raggedright\arraybackslash}p{2cm} *{3}{>{\centering\arraybackslash}p{1.5cm}} >{\centering\arraybackslash}p{2.5cm}}
      \toprule
      Parameter      & EqEq  & EqOv  & OvOv  & OvOv (\textbf{pomp})$^*$ \\
      \midrule
      \(\beta_1\)    & 12.15 & 12.74 & 11.48 & $\Sexpr{round(est_mle_vec["beta1"],2)}$\\
      \(\beta_2\)    & 0.22  & 0.21  & 0.25  & $\Sexpr{round(est_mle_vec["beta2"],2)}$\\
      \(\beta_3\)    & 0.34  & 0.31  & 0.35  & $\Sexpr{round(est_mle_vec["beta3"],2)}$\\
      \(\phi\)       & 0.017 & 0.14  & 0.14  & $\Sexpr{round(est_mle_vec["phi"],2)}$\\
      \(\rho\)       & 0.022 & 0.19  & 0.16  & $\Sexpr{round(est_mle_vec["beta11"],2)}$\\
      \(\sigma^2_q\) & n/a   & 0.042 & 0.021 & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$\\
      \(\sigma_\xi\) & n/a   & n/a   & 66.89 & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
      \addlinespace
      \multicolumn{1}{l}{\(S_{10}\)} & \multicolumn{3}{c}{$3876549^*$} & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$\\
      \multicolumn{1}{l}{\(I_{10}\)} & \multicolumn{3}{c}{$30351$}     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$\\
      \multicolumn{1}{l}{\(R_{10}\)} & \multicolumn{3}{c}{$1315221$}   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$\\
      \multicolumn{1}{l}{\(S_{20}\)} & \multicolumn{3}{c}{$57139612$}  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$\\
      \multicolumn{1}{l}{\(I_{20}\)} & \multicolumn{3}{c}{$871$}       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$\\
      \multicolumn{1}{l}{\(R_{20}\)} & \multicolumn{3}{c}{$302852$}    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$\\
      \multicolumn{1}{l}{\(S_{30}\)} & \multicolumn{3}{c}{$19573727$}  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$\\
      \multicolumn{1}{l}{\(I_{30}\)} & \multicolumn{3}{c}{$2550$}      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$\\
      \multicolumn{1}{l}{\(R_{30}\)} & \multicolumn{3}{c}{$131092$}    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$\\
      \midrule
      \addlinespace
      \textbf{AIC}  & $98866.65$ & $15154.75$ & $13778.08$ & $\Sexpr{round((-2)*((length(wwr_mle) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$\\
      \addlinespace
      \bottomrule
    \end{tabular}
    \begin{tablenotes}
      \small
      \item[*] The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be \textbf{fixed} in \cite{wwr}. The results of the estimates of the OvOv (\textbf{pomp}) model are obtained by using the Iterated Filtering algorithm with 3 rounds and 100 iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round. 
    \end{tablenotes}
  \end{threeparttable}
\end{table}
